---
title: "A Machine Profile Summary"
author:
  - John C. Nash, University of Ottawa, Canada
date: "08/06/2021"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Providing a characterization of a particular computing environment for R calculations

If we want to compare results computed on different machines, we need a way to provide 
measures and identifiers of the particular computing environment at hand. This document
attempts to summarize some possibilities available in the R statistical software and language.

# Some possible information desired

There are multiple ways to get information of the type useful for characterizing the
computing infrastructure used for a particular test or timing. We will need to
ensure that identifiers of particular pieces of information are unique, and that 
common identifiers really do specify identical information. Otherwise, we need
to develop new names for our set of information.

## Temporal information

Most computatins should have a time/date stamp that is unlikely to be confused with any
other, at least in combination with a machine name and/or other tags.

?? tdstamp

## Identifier for the computing environment

- machine name
- operating system and version
- compiler or interpreter version
- specific variations, if any, that should be noted


## Hardware information

- cpu
- cores
- operating cycle speed??
- RAM total
- RAM available, possibly indicating swap space and usage
- RAM speed (could be variable)

## Software information

?? what to put in.

## R tools for machine information

### sessionInfo()

```{r sessionInfox}
si <- sessionInfo()
si <- as.vector(si)
si
```

### benchmarkme

- uses `proc.time()` for timings
- has single core and parallel timings
- ?? does it do GPUs
- provides far too much for most needs i.e., a succinct summary of
  how capable a computing system should be

```{r benchmarkmex}
library(benchmarkmeData)
library(benchmarkme)
ls(package:benchmarkmeData)
ls(package:benchmarkme)
lsf.str("package:benchmarkme")
lsf.str("package:benchmarkmeData")
## This next line takes some time
benchmark_std()

get_byte_compiler()
get_linear_algebra()
get_sys_details()
```



### calceps.R

This is a port of Mike Malcolm's ENVRON to R. ??ref. It computes the 
machine precision, radix and number of radix digits. 

?? get file into chunk

```
calceps.R


- calceps.R example

### Sys.info()

```{r sysinfox}
Sys.info()
```



## Issues relateing to RNG

## Issues relating to compilation of R 

- LAPACK
- BLAS
- others?


# Tools for accessing and clearing environments and dataframes

?? should this be here -- probably to list things in the workspace

```
sys.frame()
sys.frames()
sys.status()
sys.on.exit()
sys.parents()
sys.calls()
```

??JN exp(sin(cos())) benchmark
