---
title: "Understanding nlsModel in the base R nls() function"
author:
  - Arkajyoti Bhattacharjee, Indian Institute of Technology, Kanpur
  - John C. Nash, University of Ottawa, Canada
date: "05/06/2021"
output: pdf_document
bibliography: ImproveNLS.bib
---

<!-- - Heather Turner, University of Warwick, UK -->

# Objective

The R function `nls()` for nonlinear least squares modeling uses a structure where
a "model" object named `m` is built from an expression in the `nls` call. This 
object contains a great deal of functionality in the form of data and functions that
different parts of the `nls` infrastructure can use. In particular, `m` contains
functions to determine if the estimation of the model is "converged" (actually if
the process is "terminated") as well as computations of residuals and the search
direction for the next set of parameters. 

This article is an attempt to document and test the features of the nlsModel object `m`.

# Location of the nlsModel code

The code is found in the file `./R-devel/src/library/stats/R/nls.R` 

Note that there is a version `nlsModel.plinear` for use with partially linear models
(?? a ref would be helpful here). The default case and the use of bounds constraints
using the `algorithm` setting "port" are handled by `nlsModel`. 

Here we will at time of writing (June 2021) only deal with the default case.

# Generating an 'm`' object

Let us use the Croucher problem and/or the Hobbs weed infestation problem (@jncnm79, page 120) again.


```{r croucherex}
# Croucher-example1.R -- https://walkingrandomly.com/?p=5254
# construct the data vectors using c()
xdata = c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
ydata = c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)

# some starting values
p1 = 1
p2 = 0.2

Cdata<-data.frame(xdata, ydata)

# do the fit
## Try numericDeriv() function
Cform <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata)
Ctheta<-c("p1","p2")


## nmodc1<-nlspkg::nlsModel(form=Cform, data=Cdata, start=c(p1,p2), wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
## print(str(nmodc1))

```

```{r hobbsex}
# Data for Hobbs problem
ydat  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
            38.558, 50.156, 62.948, 75.995, 91.972) # for testing
tdat  <-  seq_along(ydat) # for testing

# A simple starting vector -- must have named parameters for nlxb, nls, wrapnlsr.
start1  <-  c(b1=1, b2=1, b3=1)
eunsc  <-   y ~ b1/(1+b2*exp(-b3*tt))
str(eunsc)
# Can we convert a string form of this "model" to a formula
ceunsc <- " y ~ b1/(1+b2*exp(-b3*tt))"
str(ceunsc)

weeddata1  <-  data.frame(y=ydat, tt=tdat)

## Now ready to try things out.
library(nlsalt) # ?? needed because base R does not export nlsModel()
nmod1<-nlsModel(form=eunsc, data=weeddata1, start=start1, wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
str(nmod1)
ls.str(nmod1)
print(nmod1)
```

