---
title: "Understaning nlsModel in the base R nls() function"
author:
  - Arkajyoti Bhattacharjee, Indian Institute of Technology, Kanpur
  - John C. Nash, University of Ottawa, Canada
date: "05/06/2021"
output: pdf_document
bibliography: ImproveNLS.bib
---

<!-- - Heather Turner, University of Warwick, UK -->

```{r setup, include=FALSE}
rm(list=ls()) # clear the workspace for this document
knitr::opts_chunk$set(echo = TRUE)
## setup for this document
library(microbenchmark)  # timing

printsum <- function(xx){ print(summary(xx))} # May be needed
traceval  <-  TRUE  # traceval set TRUE to debug or give full history
#  Set to FALSE when we don't need extensive output
```

# nlsModel()

<!-- - created script `tnlsModel.R` as first step to create and produce an "m" object -->
<!--   (In Used-PossAbandon as of July 12) -->

- Current understanding (JN): nlsModel (and nlsModel.plinear too probably) creates
  an R object that it labels as class "nlsModel". This object contains functions
  that are called from nls.c::nls_iter to run the interation and estimate the 
  parameters in the model. There seem to be some extraneous functions, and we can
  hopefully learn enough to remove the extras.
  
- The current structure is to particularize the functions in "m" so the (essentially)
  external nls.c code acts on these. As a first goal, and part of learning how things
  work, we will want to replace the nls.c::nls_iter with all-R equivalent. 
  
# A script to examine the output of nlsModel()

We will use the Hobbs weed infestation problem (@jncnm79, page 120) again.
?? AB: we should build a set of test problems that are easy to try out. Let us
discuss the examples that are in the help for nls. (?nls in R will show them.)

```{r hobbsex}
# Data for Hobbs problem
ydat  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
            38.558, 50.156, 62.948, 75.995, 91.972) # for testing
tdat  <-  seq_along(ydat) # for testing

# A simple starting vector -- must have named parameters for nlxb, nls, wrapnlsr.
start1  <-  c(b1=1, b2=1, b3=1)
eunsc  <-   y ~ b1/(1+b2*exp(-b3*tt))
str(eunsc)
# Can we convert a string form of this "model" to a formula
ceunsc <- " y ~ b1/(1+b2*exp(-b3*tt))"
str(ceunsc)

weeddata1  <-  data.frame(y=ydat, tt=tdat)

## Now ready to try things out.
library(nlspkg) # ?? needed because base R does not export nlsModel()
nmod1<-nlsModel(form=eunsc, data=weeddata1, start=start1, wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
str(nmod1)
ls.str(nmod1)
print(nmod1)
```

# How contents of the "m" object are used

This section refers to `nls()` in baseR. Replacement functions will likely be different.

# References


