---
title: "Understanding nlsModel in the base R nls() function"
author:
  - Arkajyoti Bhattacharjee, Indian Institute of Technology, Kanpur
  - John C. Nash, University of Ottawa, Canada
date: "05/06/2021"
output: pdf_document
bibliography: ImproveNLS.bib
---

<!-- - Heather Turner, University of Warwick, UK -->

# Objective

The R function `nls()` for nonlinear least squares modeling uses a structure where
a "model" object named `m` is built from an expression in the `nls` call. This 
object contains a great deal of functionality in the form of data and functions that
different parts of the `nls` infrastructure can use. In particular, `m` contains
functions to determine if the estimation of the model is "converged" (actually if
the process is "terminated") as well as computations of residuals and the search
direction for the next set of parameters. 

This article is an attempt to document and test the features of the nlsModel object `m`.

# ISSUES

- Why is there only `upper` in the arguments of nlsModel(), i.e., no `lower`? 
  For the default algorithm, upper makes no sense either.

- Why do we have noVarying versions of some functions? And with 
  so many attributes? But also the function setVarying?
  Is this possibly a way to specify the parameters that are 
  "free" in the Nash and Walker-Smith sense of bounds and
  masks?
  
- Where is the QR decomposition actually executed? This is
  key to establishing the Gauss-Newton iteration.
  
# Location of the nlsModel code

The code is found in the file `./R-devel/src/library/stats/R/nls.R` 

Note that there is a version `nlsModel.plinear` for use with partially linear models
(?? a ref would be helpful here). The default case and the use of bounds constraints
using the `algorithm` setting "port" are handled by `nlsModel`. 

Here we will at time of writing (June 2021) only deal with the default case.

# Generating an 'm`' object

Let us use the Croucher problem and/or the Hobbs weed infestation problem (@jncnm79, page 120) again.


```{r croucherex}
# Croucher-example1.R -- https://walkingrandomly.com/?p=5254
# construct the data vectors using c()
xdata = c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
ydata = c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)

# some starting values
p1 = 1
p2 = 0.2

Cdata<-data.frame(xdata, ydata)

# do the fit
## Try numericDeriv() function
Cform <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata)
Ctheta<-c("p1","p2")
Cstart<-c(p1=p1, p2=p2)


nmodc1<-nlsalt::nlsModel(form=Cform, data=Cdata, start=Cstart, wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
print(str(nmodc1))

# nmodc1a<-nlsalt::nlsModel(form=Cform, data=Cdata, start=Cstart, wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
# print(str(nmodc1a))

```

```{r hobbsex}
# Data for Hobbs problem
ydat  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
            38.558, 50.156, 62.948, 75.995, 91.972) # for testing
tdat  <-  seq_along(ydat) # for testing

# A simple starting vector -- must have named parameters for nlxb, nls, wrapnlsr.
start1  <-  c(b1=1, b2=1, b3=1)
eunsc  <-   y ~ b1/(1+b2*exp(-b3*tt))
str(eunsc)
# Can we convert a string form of this "model" to a formula
ceunsc <- " y ~ b1/(1+b2*exp(-b3*tt))"
str(ceunsc)

weeddata1  <-  data.frame(y=ydat, tt=tdat)

## Now ready to try things out.
library(nlsalt) # ?? needed because base R does not export nlsModel()
nmodh1<-nlsalt::nlsModel(form=eunsc, data=weeddata1, start=start1, wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
str(nmodh1)
ls.str(nmodh1)
print(nmodh1)
```

# Examining the nlsModel output

The output of `nlsModel` is an object of class attribute "nlsModel" with 16 elements. (??Is this always the case, 
or can the call cause elements to be added or not produced?)

## resid()

This function computes the residuals in the current environment. (?? Is this .GlobalEnv? Can
that environment be changed i.e., is it an argument to resid()?)

Note that these are WEIGHTED residuals. As at 2021-7-13, nlsj::nlsj() returns zero valued
elements for those data points excluded via argument `subset`, but the Jacobian is full
sized. Do we want to apply zero weights to reduce the dimensions to make things "the same"??

```{r tresid}
cat("Display Croucher residuals from nmodc1 as currently set:\n")
print(nmodc1$resid())
cat("Display Hobbs residuals from nmodc1 as currently set:\n")
print(nmodh1$resid())
```


## fitted()

This function computes the fitted values of the nonlinear model in the current environment.
The output includes an attribute, "gradient", which is actually the Jacobian matrix of which
the k, j element is the 

```{r fitted}
cat("Display Croucher fitted model values from nmodc1 as currently set:\n")
print(nmodc1$fitted())
cat("Display Hobbs fitted model values from nmodc1 as currently set:\n")
print(nmodh1$fitted())
```


##  formula()  

```{r tformula}  
cat("Display Croucher formula from nmodc1 as currently set:\n")
print(nmodc1$formula())
cat("Display Hobbs formula from nmodc1 as currently set:\n")
print(nmodh1$formula())


```



##  deviance()  

```{r tdeviance}  
cat("Display Croucher deviance from nmodc1 as currently set:\n")
print(nmodc1$deviance())
cat("Display Hobbs deviance from nmodc1 as currently set:\n")
print(nmodh1$deviance())
```

##  lhs()  

```{r tlhs}  
cat("Display Croucher lhs from nmodc1 as currently set:\n")
print(nmodc1$lhs())
cat("Display Hobbs lhs from nmodc1 as currently set:\n")
print(nmodh1$lhs())
```

##  gradient()  

```{r tgradient}  
cat("Display Croucher gradient from nmodc1 as currently set:\n")
print(nmodc1$gradient())
cat("Display Hobbs gradient from nmodc1 as currently set:\n")
print(nmodh1$gradient())
```

##  conv()  

```{r tconv}
cat("Display Croucher conv from nmodc1 as currently set:\n")
print(nmodc1$conv())
cat("Display Hobbs conv from nmodc1 as currently set:\n")
print(nmodh1$conv())

```

##  incr()  

?? This function appears to return the current search direction from the Gauss-Newton
equations. We need to be very CAREFUL as we augment the `nls()` procedure to allow for
Levenberg-Marquardt stabilizations and some different approaches to the solution
iteration.

```{r tincr}  
cat("Display Croucher incr from nmodc1 as currently set:\n")
print(nmodc1$incr())
cat("Display Hobbs incr from nmodc1 as currently set:\n")
print(nmodh1$incr())
```

##  setVarying(vary = rep_len(TRUE, np))

?? This seems to return the number of free (i.e., unconstrained) parameters in the
current nonlinear model. 


```{r tsetVarying(vary = rep_len(TRUE, np)) }
cat("Display Croucher setVarying from nmodc1 as currently set:\n")
print(nmodc1$setVarying())
cat("Display Hobbs setVarying from nmodc1 as currently set:\n")
print(nmodh1$setVarying())
```

##  setPars(newPars)  

?? This function does not seem to execute. Listing it shows some quite 
involved operations on the weghts, the model, the deviance and the QR
decomposition of the Jacobian.

```{r tsetPars(newPars), eval=FALSE}
newPars <- c(p1=0.95, p2=0.15)
cat("Display Croucher setPars(newPars) from nmodc1 as currently set:\n")
print(nmodc1$setPars(newPars)())
cat("Display Hobbs setPars(newPars) from nmodc1 as currently set:\n")
newPars<- c(b1=1.1, b2=1.4, b3=0.1)
print(nmodh1$setPars(newPars)())
```

##  getPars()  

This function and getAllPars() are defined identically in nlsModel() as

```
	     getPars = function() getPars(),
	     getAllPars = function() getPars(),
```

within the assignment of the list elements in creating the "m" object at the end
of `nlsModel()`.

Prior to this, we have 


```{r tgetPars}  
cat("Display Croucher getPars from nmodc1 as currently set:\n")
print(nmodc1$fitted())
cat("Display Hobbs getPars from nmodc1 as currently set:\n")
print(nmodh1$getPars())
```

##  getAllPars()

This function returns all parameters for the current nonlinear model as vector of
named quantities. It is NOT a list as shown in the example.

```{r tgetAllPars}  
cat("Display Croucher getAllPars from nmodc1 as currently set:\n")
print(str(nmodc1$getAllPars()))
print(nmodc1$getAllPars())
is.list(nmodc1$getAllPars())
is.numeric(nmodc1$getAllPars())
cat("Display Hobbs getAllPars from nmodc1 as currently set:\n")
print(nmodh1$getAllPars())
```

##  getEnv() 

The `getEnv()` function returns the identifier of the current environment for the 
nonlinear model in focus. By using the `ls()` function we can display the names of
the objects available in this environment.

```{r tgetEnv}
cat("Display Croucher fitted model values from nmodc1 as currently set:\n")
print(nmodc1$getEnv())
cat("ls(nmodc1$getEnv():")
print(ls(nmodc1$getEnv()))
cat("Display Hobbs fitted model values from nmodc1 as currently set:\n")
print(nmodh1$getEnv())
print(ls(nmodh1$getEnv()))
```


##  trace() 

The "trace" function seems to display information about the current state of the model estimation.

ISSUE: because there is a "trace" argument to nls() (and to functions in package `nlsr` and others),
would it not be better to use `tracefn()`??

Output from the `trace()` appears to be the current evaluated `deviance` (essentially the
residual sum of squares), then in brackets the value of the `convCrit()` function -- the 
value that is compared to a convergence tolerance, and then the current parameters.
In printing the evaluated function, NULL is appended. Is this because no return value is
provided??

```{r ttrace} 
cat("Display Croucher trace from nmodc1 as currently set:\n")
print(nmodc1$trace())
cat("Display Hobbs trace from nmodc1 as currently set:\n")
print(nmodh1$trace())
```

##  Rmat()  

"Rmat" appears to return the current R matrix of the QR decomposition.

This could lead to some confusion if we build an augmented matrix for the
Levenberg-Marquardt and LM + Nash variant of the stabilization. We may need to
watch this carefully to see where it is used. A search shows ONLY
usage within the packages `nlspkg`  and `nlsalt` to compute `XtXinv`, 
to get the variance-covariance of the parameters in `summary.nls()`.

?? We need to decide how to generate `Rmat` for the purposes of getting
this information. In particular, in running the estimation iteration, we
augment the matrix, so it is important to think how to get the right 
information for `summary.nls`.


```{r tRmat}  
cat("Display Croucher Rmat from nmodc1 as currently set:\n")
print(nmodc1$Rmat())
cat("Display Hobbs Rmat from nmodc1 as currently set:\n")
print(nmodh1$Rmat())
```

##  predict(newdata = list(), qr = FALSE)  

This function allows predicted values of the model to be calculated for dependent variable
data in the list "newdata". 

?? What does qr=FALSE do??

```{r tpredict}  
cat("Display Croucher predict from nmodc1 as currently set:\n")
cat("first with no new parameters -- we get the fitted values\n")
print(nmodc1$predict())
cat("Now with some new data\n")
newcdata<-data.frame(xdata=0.1, ydata=1.1)
print(nmodc1$predict(newdata=newcdata))
cat("Display Hobbs predict from nmodc1 as currently set:\n")
cat("first with no new parameters -- we get the fitted values\n")
print(nmodh1$predict())
cat("Now with some new data\n")
newhdata<-data.frame(tt=c(13,14), y=c(120,200))
print(nmodh1$predict(newdata=newhdata))

cat("trying again with qr=TRUE\n")
print(nmodh1$predict(newdata=newhdata, qr=TRUE))
```

Using "qr=TRUE" does not seem to make any difference. Need to delve further??

# Attempt to understand nlsModel in detail

A new file `nlsModelx.R` was created that establishes the `nlsModelx()` function.

```{r modx1}
nmodc1<-nlsalt::nlsModelx(form=Cform, data=Cdata, start=Cstart, wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
print(str(nmodc1))

nmodh1<-nlsalt::nlsModelx(form=eunsc, data=weeddata1, start=start1, wts=NULL, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
str(nmodh1)
```

# Discussion

Here we will look at some of the issues of using nlsModel. In particular, we will examine
how this function may need to be modified to make it easier to maintain and use.

# References
